The Ontological Modal Collapse \\ as a Collapse of the Square of Opposition

### 2016-03-25 Papers/2014/Universalis/ModalCollapse.tex
some improvements for final version of modal collapse paper

G\"{o}del defines God as a being who possesses all \emph{positive}
properties and states a few reasonable (but debatable) axioms that
such properties should satisfy.  
The overall idea of G{\"o}del's proof
is in the tradition of Anselm's argument, who defined God as some
entity of which nothing greater can be conceived. 
Anselm argued that
existence in the actual world would make 
such an assumed being even
greater (more perfect); 
hence, by definition, God must exist. However,
for Anselm existence was treated as a predicate and 
the possibility of God's existence was assumed as granted. 
These issues were criticized by Kant and Leibniz, 
respectively, and they were addressed in the work of G\"odel.

Nevertheless, G{\"o}del's work still leaves room for criticism.  In
particular, his axioms are so strong that, when assuming unrestricted
comprehension principles\footnote{A possible direction to remedy modal
  collapse, as studied e.g. by Koons
  \cite{koons06:_sobel_goedel_ontol_proof} is to impose restrictions
  on the domain of properties.}, 
%\marginpar{added footnote}
they entail a \emph{modal collapse}
\cite{Sobel1987,sobel2004logic}: everything that is the case is so
necessarily.  There has been an impressive body of recent and ongoing
work
(cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,bjordal99,fuhrmann05:_exist_notwen,Hajek2002,Hajek2008,ContemporaryBibliography}
and the references therein) proposing solutions for the modal
collapse.  The goal of this contribution is to discuss the modal
collapse from the point of view of the modal square of opposition. 
Ontological arguments typically rely on an inversion of the normal direction of entailment
in the modal square of opposition for one particular proposition (i.e. God's existence), 
and the modal collapse shows that this inversion happens for all propositions, resulting in a total collapse of the modal square of opposition.

@@

The technique enabling these experiments is the embedding of 
quantified modal logics into higher-order logics 
\cite{J23,B9,C36}, for which automated theorem provers exist. 
This technique has already been successfully employed in the 
verification and reconstruction of G\"odel's proof 
\cite{J28,J30,W50,J29}, and a detailed 
mathematical description is available in \cite{C40}.

>>>>

G\"{o}del defines God as a being who possesses all \emph{positive}
properties and states a few reasonable (but debatable) axioms that
such properties should satisfy.  The overall idea of G{\"o}del's proof
is in the tradition of Anselm's argument, who defined God as some
entity of which nothing greater can be conceived.  Anselm argued that
existence in the actual world would make such an assumed being even
greater (more perfect), hence, by definition, God must exist. However,
for Anselm existence was treated as a predicate and the possibility of
God's existence was assumed as granted.  These issues were criticized
by Kant and Leibniz, respectively, and they were addressed in the work
of G\"odel.

Nevertheless, G{\"o}del's work still leaves room for criticism.  In
particular, his axioms are so strong that, when assuming unrestricted
comprehension principles\footnote{A possible direction to remedy modal
  collapse, as studied e.g. by Koons
  \cite{koons06:_sobel_goedel_ontol_proof} is to impose restrictions
  on the domain of properties.}, 
%\marginpar{added footnote}
they entail a \emph{modal collapse}
\cite{Sobel1987,sobel2004logic}: everything that is the case is so
necessarily.  There has been an impressive body of recent and ongoing
work
(cf.~\cite{sobel2004logic,Fitting,anderson90:_some_emend_of_goedel_ontol_proof,AndersonGettings,bjordal99,fuhrmann05:_exist_notwen,Hajek2002,Hajek2008,ContemporaryBibliography}
and the references therein) proposing solutions for the modal
collapse.  The goal of this contribution is to discuss the modal
collapse from the point of view of the modal square of opposition. 
Ontological arguments typically rely on an inversion of the normal direction of entailment
in the modal square of opposition for one particular proposition (i.e. God's existence), 
and the modal collapse shows that this inversion in fact occurs for all propositions, resulting in a total collapse of the modal square of opposition.

@@

The technique enabling these experiments is the embedding of 
quantified modal logics into higher-order logics 
\cite{J23,B9,C36}, for which automated theorem provers exist. 
This technique has already been successfully employed in the 
verification and reconstruction of G\"odel's proof 
\cite{J28,J30,J29}, and a detailed 
mathematical description is available in \cite{C40}.
